Staged Abstract Interpreters

### 2019-04-06 paper/40_absinterp.tex
Grammar absinterpreter

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time abstraction, but for the purpose of clearly
illustrating the idea, we will just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components, namely
abstract values, abstract environments and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. An simple instance of lattice we used is the power
set abstract domain (shown in Figure \ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is intersect operation.
Other lattices used in the remainder of the paper, such as products and maps, can be
implemented similarly, or lifted from existing lattices element-wise or
point-wise. Naturally, abstract domains such intervals also can be implemented
in a stage polymorphic way.

@@

We follow the \textit{abstracting definitional interpreters} (ADIs) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter. We first widen the concrete values to @AbsValue@: numbers
are lifted to a singleton object @IntTop@ representing the set of all numbers;
closures remain the same. Then the type @Value@ is sets of @AbsValue@ (we abuse
the type names a little bit, as the generic interface defined the abstract type
@Value@).
The address space is constrained to be finite to ensure the computatbility of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After settling the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, meaning an address may point to an over-approximated set of values
occurred at runtime.

@@

As of now, the store is a mapping from addresses to sets of abstract values.
Therefore, when applying a function, for instance @f(a)@, we may retrieve a set
of closures by dereferencing the address of @f@. We need to nondeterministically
apply all of the target closures.
Another nondeterminism happens when evaluating conditionals, we have to evaluate
both branches if the condition can not be determined. In other words, we will
have multiple computing results.

@@

This observation leads us to the type definition of @Result@: in concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a set of such pairs
@Set[(Value, Store)]@ to represent multiple results. The results is also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

\paragraph{Monads for Abstract Interpretation} Now we incorporate the
nondeterminism and caching into the monad stack. For concrete evaluation, we
already have a reader effect for environments and a state effect for stores. We
further add three more effects based on it: the nondeterminism effect is
represented by the @SetT[M[_], A]@ monad transformer, inside of which, we have
a @ReaderT@ for a @in@ cache, and a @StateT@ for an @out@ cache. \todo{highlight
where changed}

@@

The sketch implementation of @SetT@ transformer is shown below. The value @run@
encapsulated by the monad is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and
@A@ is the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[SetT[A]]@ and obtain a value of @Set[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@SetT[M, B]@.

@@

\paragraph{Primitive Operations} Some of the primitive operations are changed
according to the new monad stack. A notable change is that the store update
operation joins the value with the existing values, for the same address. Under
abstract interpretation, we may have to explore the results from both branches
of a conditional expression for soundness. In @br0@, we combine the results
using @mplus@ from @MonadPlus@, which requires the values are join-semilattices
(in our case, @Set@s and @Map@s).

@@

Figure \ref{fig:coind_cache} shows the co-inductive caching algorithm that
closes the open-recursive @eval@.
The idea is to first set up two caches called @in@ and @out@ both of type
@Cache@. Then similar to Kleene's iteration algorithm, we compute the analysis
result iteratively. The @in@ cache contains what we already have computed from
the last iteration, and the @out@ cache is what we would have computed after this
iteration joined with the contents from @in@ cache.
When starting a new iteration, @in@ is set to be the result of the last
iteration, i.e., @out@; and @out@ is assigned to be empty.

During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation, if
yes then result is directly returned through the monads.

>>>>

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time abstraction, but for the purpose of clearly
illustrating the idea, we will just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components, namely
abstract values, abstract environments, and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. A simple instance of lattice we used is the power
set abstract domain (shown in Figure \ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is the intersect operation.
Other lattices used in the remainder of the paper, such as products and maps, can be
implemented similarly or lifted from existing lattices element-wise or
point-wise. Naturally, abstract domains such intervals also can be implemented
in a stage polymorphic way.

@@

We follow the \textit{abstracting definitional interpreters} (ADIs) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter. We first widen the concrete values to @AbsValue@: numbers
are lifted to a singleton object @IntTop@ representing the set of all numbers;
closures remain the same. Then the type @Value@ is sets of @AbsValue@ (we abuse
the type names a little bit, as the generic interface defined the abstract type
@Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After settling the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, meaning an address may point to an over-approximated set of values
occurred at runtime.

@@

As of now, the store is a mapping from addresses to sets of abstract values.
Therefore, when applying a function, for instance, @f(a)@, we may retrieve a set
of closures by dereferencing the address of @f@. We need to nondeterministically
apply all of the target closures.
Another nondeterminism happens when evaluating conditionals, we have to evaluate
both branches if the condition can not be determined. In other words, we will
have multiple computing results.

@@

This observation leads us to the type definition of @Result@: in a concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a set of such pairs
@Set[(Value, Store)]@ to represent multiple results. The results are also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

\paragraph{Monads for Abstract Interpretation} Now we incorporate the
nondeterminism and caching into the monad stack. For concrete evaluation, we
already have a reader effect for environments and a state effect for stores. We
further add three more effects based on it: the nondeterminism effect is
represented by the @SetT[M[_], A]@ monad transformer, inside of which, we have
a @ReaderT@ for a @in@ cache and a @StateT@ for an @out@ cache. \todo{highlight
where changed}

@@

The sketch implementation of the @SetT@ transformer is shown below. The value @run@
encapsulated by the monad is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and
@A@ is the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[SetT[A]]@ and obtain a value of @Set[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@SetT[M, B]@.

@@

\paragraph{Primitive Operations} Some of the primitive operations are changed
according to the new monad stack. A notable change is that the store update
operation joins the value with the existing values, for the same address. Under
abstract interpretation, we may have to explore the results from both branches
of a conditional expression for soundness. In @br0@, we combine the results
using @mplus@ from @MonadPlus@, which requires the values are join-semilattices
(in our case @Set@s and @Map@s).

@@

Figure \ref{fig:coind_cache} shows the co-inductive caching algorithm that
closes the open-recursive @eval@.
The idea is to first set up two caches called @in@ and @out@ both of type
@Cache@. Then similar to Kleene's iteration algorithm, we compute the analysis
result iteratively. The @in@ cache contains what we already have computed from
the last iteration, and the @out@ cache is what we would have computed after this
iteration joined with the contents from the @in@ cache.
When starting a new iteration, @in@ is set to be the result of the last
iteration, i.e., @out@; and @out@ is assigned to be empty.

During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation, if
yes, then the result is directly returned through the monads.
